Optimal Finite Element Integration

### 2015-08-03 TOMS_Paper/assembly.tex
Rewrite the introduction

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become increasingly popular over the last years. In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language. The mathematical specification is manipulated and then passed to a form compiler, which generates a representation of local assembly operations. These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution. Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more computationally expensive, covering a significant fraction of the overall computation run-time.

Producing high performance implementations is, however, non-trivial. The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process. Traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions. Polyhedral-model-based source-to-source compilers, for instance~\cite{PLUTO}, mainly apply aggressive loop optimizations, such as tiling, but these are not particularly helpful in our context. This lack of suitable optimizing tools has led to the development of a number of higher-level approaches to maximize the performance of local assembly kernels.  In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''. \cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of finite element integration. In~\cite{Luporini}, we have made one step forward by showing that different forms, on different platforms, require distinct sets of transformations if close-to-peak performance must be reached, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality. The problem of optimizing local assembly routines has been tackled recently also for GPU architectures, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.

Our research has resulted in ....
... we build on our previous work~\cite{Luporini} ... and present a ... We argue that for complex, realistic forms, peak performance can be achieved only by  ... 
... also low-level optimisation ...
...also for the first time we provide a formal explanation of the math in terms of compiler theory
-  Deficiencies of previous approaches

This is all implemented in COFFEE, which in turn is integrated with the Firedrake framework. We provide an extensive and unprecedented performance evaluation across a number of forms of increasing complexity, including some based on complex (hyperelasticity) models. We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions. To clearly distinguish the improvement achieved by this work, we will compare, for each test case, X sets of code variants: 1) unoptimized code, i.e. a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e. the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; ....

@@

These two modes have been implemented in the Fenics Form Compiler. In this compiler, a simple heuristic is used to choose the most suitable mode for a given form. It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically~\cite{Fenics}. We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-syntesis}. For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of selecting the optimal mode at the monomials level -- no heuristics, no "global" choice of the mode for all monomials in the form -- without affecting the cost of code generation.

>>>>

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects, which have become increasingly popular over the last years. In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language. The mathematical specification is manipulated by a form compiler, which then generates a representation of assembly operators. By applying these operators to an element in the discretized domain a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed. The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, thus making assembly covering a significant fraction of the overall runtime.

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or "flops", than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific transformation systems or synthesizers. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations that a user should not be expected to write ``by hand''. In~\citeN{Kirby-FEM-opt} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. There is also an on-going effort to produce a novel form compiler, and therefore a novel code generation system for assembly operators, called UFLACS (\cite{Uflacs}).

In spite of a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? With this paper, we aim to make a first step in the direction of solving this problem.

We summarize the contributions of this research as follows
\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
\item We implemented the theory in our compiler, COFFEE\footnote{...}, which is the default optimization system in the Firedrake framework. Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
\item We provide an unprecedented (for extension) performance evaluation comparing several code transformation/synthesizers systems in a wide range of forms of increasing complexity, which is essential to demonstrate our optimality claim.
\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we deal with (i.e., optimize) them
\item We enrich the set of low-level transformations described in~\citeN{Luporini}
\end{itemize}

@@

These two modes have been implemented in the FEniCS Form Compiler. In this compiler, a simple heuristic is used to choose the most suitable mode for a given form. It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically~\cite{Fenics}. We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}. For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of selecting the optimal mode at the monomials level -- no heuristics, no "global" choice of the mode for all monomials in the form -- without affecting the cost of code generation.
